---
title: "Questions for Lab 1, 2 and 3"
author: "Leo lahti & Rajesh Shigdel"
date: "`r Sys.Date()`"
output: html_document
---


# Tutorial Aims 
In this tutorial we will learn feature selection, dimension reduction, clustering, visualization, analysis and interpretation  using  
 miaverse (mia = MIcrobiome Analysis) and miaViz - (Microbiome Analysis Plotting and Visualization) to explore patterns in human gut microbiome datasets. 

The example in this tutorial are mainly based on this online book        
[Orchestrating Microbiome Analysis](https://microbiome.github.io/OMA/)

In this tutorial we use data from [The Human Gut Microbiome Atlas (HGMA)](https://www.microbiomeatlas.org/)



# Lab 1 
   
   
## unsupervised learning: Feature selection & dimension reduction
   
1. Make the tse (TreeSummarizedExperiment) Object for the further analysis  (use the provided script to make a tse)


## Load the data 



2. Aggregate the data to Phylum level 

```{r}
tse_phylum <- agglomerateByRank(tse, rank = "Phylum")
```



3. Subset the data at the Species level with only taking bacterial Species that are greater than 10% of prevalence in the total sample


```{r}
altExp(tse,"species") <- agglomerateByRank(se, "species")
getPrevalentTaxa(tse, detection = 0, prevalence = 10/100)
prev <- getPrevalentTaxa(tse, detection = 0, prevalence = 10/100)
                         
```


4. Subset a tse object which consist only two phyla i.e Actinobacteria", "Cyanobacteria


```{r}
se_subset_by_feature <- se[rowData(tse)$phylum %in% c("Actinobacteria", "Cyanobacteria")]
se_subset_by_feature
```

 
5. Calculate relative abundances, and store the table to assays


```{r}
tse <- relAbundanceCounts(tse)

```


6.Perform a centered log-ratio transformation (clr) (i.e. mia::transformSamples)

```{r}
tse <- transformSamples(x = tse, abund_values = "counts", method = "clr", 
                       pseudocount = 1, name = "clr_transformation")
```


# Lab 2 
       
  
## unsupervised learning: clustering & visualization
                   
 1. Visualize beta-diversity using principal coordinate analysis (PCoA);based on the Bray-Curtis dissimilarities
 
```{r}
1.#PCoA for ASV-level data with Bray-Curtis

se <- relAbundanceCounts(se)

# Pick the relative abundance table
rel_abund_assay <- assays(se)$relabundance

# Calculates Bray-Curtis distances between samples. Because taxa is in
# columns, it is used to compare different samples. We transpose the
# assay to get taxa to columns
bray_curtis_dist <- vegan::vegdist(t(rel_abund_assay), method = "bray")

# PCoA
bray_curtis_pcoa <- ecodist::pco(bray_curtis_dist)

# All components could be found here: 
# bray_curtis_pcoa$vectors
# But we only need the first two to demonstrate what we can do:
bray_curtis_pcoa_df <- data.frame(pcoa1 = bray_curtis_pcoa$vectors[,1], 
                                  pcoa2 = bray_curtis_pcoa$vectors[,2])

# Create a plot
bray_curtis_plot <- ggplot(data = bray_curtis_pcoa_df, aes(x=pcoa1, y=pcoa2)) +
  geom_point() +
  labs(x = "PC1",
       y = "PC2", 
       title = "Bray-Curtis PCoA") +
  theme(title = element_text(size = 10)) # makes titles smaller

bray_curtis_plot
```
 


 2. Visualize beta-diversity using principal coordinates analysis (PCoA); with Aitchison distance (clr transformation+ Euclidean distance)
 
```{r}

tse <- tse[, !is.na(tse$Gender)]


# Does clr transformation. Pseudocount is added, because data contains zeros. 
tse <- transformCounts(tse, method = "clr", pseudocount = 1)

# Gets clr table
clr_assay <- assays(tse)$clr

# Transposes it to get taxa to columns
clr_assay <- t(clr_assay)

# Calculates Euclidean distances between samples. Because taxa is in columns,
# it is used to compare different samples.
euclidean_dist <- vegan::vegdist(clr_assay, method = "euclidean")

# Does principal coordinate analysis
euclidean_pcoa <- ecodist::pco(euclidean_dist)

# Creates a data frame from principal coordinates
euclidean_pcoa_df <- data.frame(pcoa1 = euclidean_pcoa$vectors[,1], 
                                pcoa2 = euclidean_pcoa$vectors[,2])

# Creates the plot
euclidean_plot <- ggplot(data = euclidean_pcoa_df, aes(x=pcoa1, y=pcoa2)) +
  geom_point() +
  labs(x = "PC1",
       y = "PC2",
       title = "Euclidean PCoA with CLR transformation") +
  theme(title = element_text(size = 12)) # makes titles smaller

euclidean_plot



```
 


 3. Cluster the samples using Dirichlet- multinomial mixture model 



```{r}
tse <- agglomerateByRank(tse, rank = "Phylum", agglomerateTree=TRUE)
tse_dmn <- runDMN(tse, name = "DMN", k = 1:7)
tse_dmn


#Return information on metadata that the object contains.
names(metadata(tse_dmn))
#This returns a list of DMN objects for a closer investigation.
getDMN(tse_dmn)

```

```{r}

library(miaViz)
plotDMNFit(tse_dmn, type = "laplace")

```

```{r}
#Return the model that has the best fit.
getBestDMNFit(tse_dmn, type = "laplace")
```

```{r}
dmn_group <- calculateDMNgroup(tse_dmn, variable = "Geography",  exprs_values = "counts",
                               k = 2, seed=.Machine$integer.max)

dmn_group
```



```{r}
#Mixture weights (rough measure of the cluster size).
DirichletMultinomial::mixturewt(getBestDMNFit(tse_dmn))
```

```{r}
#Samples-cluster assignment probabilities / how probable it is that sample belongs to each cluster

head(DirichletMultinomial::mixture(getBestDMNFit(tse_dmn)))
```


```{r}
#Contribution of each taxa to each component
head(DirichletMultinomial::fitted(getBestDMNFit(tse_dmn)))
```

```{r}
#Get the assignment probabilities
prob <- DirichletMultinomial::mixture(getBestDMNFit(tse_dmn))
# Add column names
colnames(prob) <- c("comp1", "comp2")

# For each row, finds column that has the highest value. Then extract the column 
# names of highest values.
vec <- colnames(prob)[max.col(prob,ties.method = "first")]

```

```{r}
# Does clr transformation. Pseudocount is added, because data contains zeros.
tse <- transformCounts(tse, method = "clr", pseudocount = 1)

# Gets clr table
clr_assay <- assays(tse)$clr

# Transposes it to get taxa to columns
clr_assay <- t(clr_assay)

# Calculates Euclidean distances between samples. Because taxa is in columns,
# it is used to compare different samples.
euclidean_dist <- vegan::vegdist(clr_assay, method = "euclidean")

# Does principal coordinate analysis
euclidean_pcoa <- ecodist::pco(euclidean_dist)

# Creates a data frame from principal coordinates
euclidean_pcoa_df <- data.frame(pcoa1 = euclidean_pcoa$vectors[,1], 
                                pcoa2 = euclidean_pcoa$vectors[,2])
```

 4. Visualize the clusters in the PCoA plot

```{r}
# Creates a data frame that contains principal coordinates and DMM information
euclidean_dmm_pcoa_df <- cbind(euclidean_pcoa_df,
                               dmm_component = vec)
# Creates a plot
euclidean_dmm_plot <- ggplot(data = euclidean_dmm_pcoa_df, 
                             aes(x=pcoa1, y=pcoa2,
                                 color = dmm_component)) +
  geom_point() +
  labs(x = "Coordinate 1",
       y = "Coordinate 2",
       title = "PCoA with Aitchison distances") +  
  theme(title = element_text(size = 12)) # makes titles smaller

euclidean_dmm_plot
```






# Lab 3 
         
                
## unsupervised learning:Analysis and interpretation 

 1. What taxa are driving the axis? Calculate the spearman correlation between PC1 and the relative abundance of the bacteria and visualize the results in a     bar plot.


 2. Visualize the dominant taxonomic group of each sample by colour on PCoA plot 

 3. Visualize gender by colour on PCoA plot 




